---
title: "TP: Anova analysis on wheat"
author: "Gabriele Degola"
date: "9/12/2021"
output:
  html_document:
    df_print: paged
  pdf_document:
    includes:
      in_header: "../utils/wrap-code.tex"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, message=FALSE, warning=FALSE}
library(tidyverse)
```

## Import the data

```{r}
ble <- read.table("../data/ble.txt", header=TRUE, sep=";", dec=".")
head(ble)
```

## Perform a 1-factor ANOVA

```{r fig.show="hold", out.width="50%"}
ggplot(ble, aes(x=variete, y=rdt)) + geom_boxplot() +
  ggtitle("Whisker boxes") + xlab("Wheat variety") + ylab("Yield")

ggplot(ble, aes(x=phyto, y=rdt)) + geom_boxplot() +
  ggtitle("Boxplot") + xlab("Phytosanitary treatment") + ylab("Yield")
```

### Anova test on wheat variety

```{r}
anova_variete <- lm(rdt ~ variete, data=ble)
summary(anova_variete)
```

p-value is very small, there is an actual influence of the wheat variety on the produced yield. We thereofre reject the null hypothesis that variety has no impact of yield. p-values of each variety are with respect of the first modality. Indeed, there is no significative difference in yield between variety 1 and 2, while 3 and 4 behave differently from variety 1.

```{r}
anova(anova_variete)
```

### Anova test on pesticide

```{r}
anova_phyto <- lm(rdt ~ phyto, data=ble)
summary(anova_phyto)
```

Big p-value, we do not reject the hypothesis that use of pesticide have no influence on the produced yield. No significative difference between the yield produced with or without pesticides.

```{r}
anova(anova_phyto)
```

Tests are also significative (good results) as data are balanced (20 entries for each variety, half with and half without pesticides).

## Perform a 2-factor ANOVA

```{r}
anova_variete_phyto <- lm(rdt ~ variete * phyto, data=ble)
summary(anova_variete_phyto)
```

```{r}
anova(anova_variete_phyto)
```

Big p-values for `phyto` and `variete:pytho`: use of pesticide do not have any impact on yield, as well as the interaction between use of pesticides and variety.
