---
title: "Lecture 3: intro to R and dplyr"
author: "Gabriele Degola"
date: "October, 2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction to R

Load and explore data:

```{r}
mtcars
```

```{r}
str(mtcars)
```

Columns of the dataframe can be selected and manipulated easily:

```{r}
mtcars$drat
```

```{r}
mtcars$drat + 1
```

```{r}
mtcars$drat > 3.5
```

Filter observations according to the value of a feature (mind the comma at the end):

```{r}
mtcars[mtcars$drat > 3.5,]
```

```{r}
mtcars[mtcars$drat > 3.5 & mtcars$cyl == 6,]
```

```{r}
mtcars[mtcars$drat > 3.5,]$carb
```

Compute feature statistics:

```{r}
mean(mtcars$mpg); sd(mtcars$mpg)
```

## dplyr

```{r, warning=FALSE, message=FALSE}
library(dplyr)
```

Use `summarise` to create a new dataframe summarising some variables:

```{r}
mtcars %>% summarise(mpg_mean = mean(mpg), mpg_sd = sd(mpg))
```

Use `group_by` to create different group according to the values of one or more variables:

```{r}
mtcars %>% group_by(cyl) %>% summarise(mpg_mean = mean(mpg), mpg_sd = sd(mpg))
```

```{r, message=FALSE}
mtcars %>% group_by(cyl, am) %>% summarise(mpg_mean = mean(mpg), mpg_sd = sd(mpg))
```

```{r, message=FALSE}
mtcars %>% group_by(cyl, carb) %>% summarise(mpg_mean = mean(mpg), mpg_sd = sd(mpg))
```

Some standard deviation values are `NaN`. Why? Because there is a single observation having certains values for `cyl` and `carb` (for example, 6 and 6).

```{r, message=FALSE}
mtcars %>% group_by(cyl, carb) %>% summarise(mpg_mean = mean(mpg), mpg_sd = sd(mpg), num = n())
```

Note that `n` is a special function to be used in a data context. `nrow` returns the number of observations in a dataframe.

```{r}
nrow(mtcars)
```

`ungroup` is used to remove grouping and perform other operations on the whole dataframe. By itself, it does not do anything:

```{r, message=FALSE}
mtcars %>% group_by(cyl, carb) %>% summarise(mpg_mean = mean(mpg), mpg_sd = sd(mpg), num = n()) %>% ungroup()
```

`mutate` adds new features, while preserving existing ones:

```{r, message=FALSE}
mtcars %>% group_by(cyl, carb) %>% summarise(mpg_mean = mean(mpg), mpg_sd = sd(mpg), num = n()) %>% ungroup() %>% mutate(mpg_gmean = mean(mpg_mean))
```

```{r, message=FALSE}
mtcars %>% group_by(cyl, carb) %>% summarise(mpg_mean = mean(mpg), mpg_sd = sd(mpg), num = n()) %>% ungroup() %>% mutate(mpg_gmean = mean(mpg_mean), deviation = mpg_mean - mpg_gmean)
```


## Plotting in R

```{r}
cars
```

```{r}
plot(cars)
```

```{r}
plot(cars$speed, cars$dist)
```
